State of the Practice for Lattice Boltzmann Method Software

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 6

As part of filling in the measurement template, we listed the artifacts for each
package, and we then categorized them by frequency. We group the artifacts into
common, uncommon, and rare categories in Table~\ref{artifactspresent}. Common
artifacts are in 16 to 24 ($>$63\%) of the software packages. Uncommon artifacts
are in 8 to 15 (30-63\%) and rare artifacts are in 1 to 7 ($<$30\%) of the
packages.  We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that
we observed in LBM repositories to those observed and recommended for research
software in general. Our comparison may point out areas where some LBM software
packages fall short of current best practices. This is not intended to be a
criticism of any existing packages, especially since in practice not every
project needs to achieve the highest possible quality. However, rather than
delve into the nuances of which software can justify compromising which
practices, we will write our comparison under the ideal assumption that every
project has sufficient resources to match best practices.

@@

In Table~\ref{Tbl_Guidelines}, each row corresponds to an artifact.  For a given
row, a checkmark in one of the columns means that the corresponding guideline
recommends this artifact.  The last column shows whether the artifact appears in
the measured set of LBM software, either not at all (blank), commonly (C),
uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  The challenge
even exists for the ubiquitous README file.  As illustrated by
\citet{PranaEtAl2018}, the content of README files shows significant variation
between projects.  Although some content is reasonably consistent, with 97\% of
README files containing at least one section describing the `What` of the
repository and 88.5 \% offering some `How` content, other categories are more
variable.  For instance, information on `Contribution`, `Why`, and `Who`,
appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
\citep{PranaEtAl2018}.

The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
popularity of recommending a given artifact, but it does not imply that the most
commonly recommended artifacts are the most important artifacts. Just because a
guideline does not explicitly recommend an artifact does not mean that the
artifact has no value.  The guideline authors may have excluded it because it is
out of the scope of their recommendations or outside their experience. For
instance, an artifact related to uninstall is only explicitly mentioned by
\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
value.  They may simply feel that uninstall is implied by install, or they may
have never asked themselves about the need for separate uninstall instructions.

@@

The top four packages (ESPREsSo, Ludwig, Palabos, and OpenLB) have all the
common artifacts, except Palabos does not have theory notes and only three
(ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
system.  The top four packages also have most of the uncommon artifacts. At the
time of data collection, only one of the four packages (Palabos) did not have a
user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. The developers later fixed this
broken link, but this is not reflected in our data because it was not present at
the time of data collection (mid-2020). Despite the broken link, Palabos does
have a detailed and informative website. As far as we could observe, Ludwig does
not appear to have a publicly visible design and OpenLB does not appear to use a
version control system. However, OpenLB's use of version numbers suggests that
its developers may privately use version control.

The top four ranked packages do not have many of the rare artifacts. None of the
top four packages have any explicit API documentation. Three of these packages
(ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
of them (OpenLB, Palabos) have an FAQ section or forum. One (OpenLB) has
verification and validation notes, and a video guide for the software.

Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
appear in the software development guidelines: i) list of related publications,
ii) performance information and iii) video guides.  The occurrence of the first
two items in this list are likely more a consequence of how we collected our
list of LBM artifacts, rather than being a unique practice of LBM developers. We
highlighted publications and performance information because we were explicitly
looking for both as part of filling in our measurement template. Related
publications and performance information certainly appear in other research
software projects; the practices just are not highlighted in guidelines because
they are implicit in other documentation recommendations, such as theory
documentation and user guides. The observation of a video guide artifact is more
novel.  \citet{Fogel2005} recommends videos for open source projects, and other
projects likely use them, but the practice has apparently not yet made it into
research software community's guidelines.  Putting aside the debate of the
effectiveness of learning from print versus video, video documentation is
growing in popularity, with a majority of Generation Z learners preferring video
to print \citep{Genota2018}.

As Table~\ref{Tbl_Guidelines} shows, the LBM community participates in most of
the practices we found listed in the general research software guidelines;
however, we did not observe, or rarely observed, some recommended practices. For
instance, we rarely saw API documentation for LBM software, but it is a
frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
vanGompelEtAl2016, OrvizEtAl2017, SSI2022, Zadka2018}.  In addition to the items
in the last column of Table~\ref{artifactspresent}, we can add the following
community recommended items that we rarely, if ever, observed:

@@

\item \textbf{A roadmap} is considered by some to be a critical document for an
organization to map out the vision and direction for a product offering
\citep{MunchEtAl2019}.  Roadmaps take different forms, but they all aim to
address the following: \begin{inparaenum}[i)]
	\item Where are we now?,
	\item Where are we going?, and
	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}.
Developers can create a roadmap by the following steps:
\begin{inparaenum}[i)]
	\item define and outline a strategic mission and product vision, 
	\item scan the environment, 
	\item revise and distill the product vision to write the product roadmap, and 
	\item estimate the product life cycle and evaluate the mix of planned
development efforts. \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
roadmap has the benefits of providing continuity of purpose, facilitating
stakeholder collaboration and assisting with prioritization \citep{Pichler2012}.
Although we saw some elements of a roadmap in the LBM projects, we did not
observe a specific artifact devoted to this purpose.

\item \textbf{A code of conduct} documents expectations and standards of ethical
behaviour and explicitly states how developers should treat one another
\citep{TouraniEtAl2017}. A code of conduct outlines rules for communication,
establishes enforcement mechanisms for violations and codifies the spirit of a
community, such that anyone can contribute comfortably regardless of, for
example, ethnicity, gender, or sexual orientation \citep{TouraniEtAl2017}. The
top three most popular code's of conduct from \citet{TouraniEtAl2017} are:
\href{https://ubuntu.com/community/code-of-conduct} {Ubuntu Code of Conduct},
\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
{Contributor Covenant}, and \href{https://www.djangoproject.com/conduct/} {Django
Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a code of
conduct on the participation of women. Improved inclusivity means a wider pool
of potential contributors. Moreover, a code of conduct facilitates meeting
ethical obligations for developers that abide by a code of ethics, such as the
IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of Ontario
code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did not
observe any codes of conduct, possibly because the number of developers for each
LBM project is small and because the idea has not yet occurred to any of the
participants.

@@

\item \textbf{Checklists} can be used to ensure that developers follow best
practices.  Some examples include checklists merging branches into master
\citep{Brown2015}, checklists for saving and sharing changes to the project
\citep{WilsonEtAl2016}, checklists for new and departing team members
\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
releases \citep{HerouxEtAl2008} and checklists for overall software quality
\citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
managing releases, but we did not find any other examples.

\item \textbf{Uninstall instructions} tell a user how to remove the software and
associated libraries from their system.  The advantage of uninstall is removing
files that consume space and potentially pollute the users build system by
causing conflicts when installing and using other software packages.  We did not
observe uninstall instructions for any of the LBM projects.

\item \textbf{API documentation} shows programmers the services or data provided
by the software application (or library) through such resources as its methods
or objects \citep{MengEtAl2018}.  API documentation is a critical factor for API
understandability \citep{MengEtAl2018} (Section~\ref{softwarequalities}). Some
API documentation can be generated using tools like Doxygen, but when learning a
new API developers mostly use code examples and tutorials \citep{MengEtAl2018}.
API documentation was rare within our sample of LBM packages.

@@

Where data is available, LBM software artifacts can be compared to other
research software domains.  Compared to ocean modelling software
\citep{JungEtAl2022}, LBM software pays more attention to artifacts related to
testing.  Compared to Medical Imaging Software \citep{Dong2021}, the artifact
frequency is similar, except LBM software is more likely to have developer
related artifacts, like Contributing, Dependency list, and Design documentation.

Table~\ref{Tbl_Guidelines} shows that many LBM projects fall short of
recommended best practices for research software.  However, LBM software is not
alone in this.  Many, if not most, research projects fall short of best
practices.   A gap exists in scientific computing development practices and
software engineering recommendations \citep{Storer2017, Kelly2007,
OwojaiyeEtAl2021_CSE}. \citet{JohansonAndHasselbring2018} observe that the
state-of-the-practice for SCS in industry and academia does not incorporate
state-of-the-art SE tools and methods.  This causes sustainability and
reliability problems \citep{FaulkEtAl2009}. Rather than benefit from capturing
and reusing previous knowledge, projects waste time and energy ``reinventing the
wheel'' \citep{deSouzaEtAl2019}.

Software requirements documentation provides a good example of the common
deviation between recommendations and practice.  Although some guidelines
recommend requirements documentation \citep{TobiasEtAl2018, HerouxEtAl2008,
SmithAndKoothoor2016}, in practice research software developers often do not
produce a proper requirements specification \citep{HeatonAndCarver2015}.
\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
found that none of the scientists created requirements specifications, unless
regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
showed requirements are the least commonly produced type of documentation for
research software in general. When looking at the pain points for research
software developers, \citet{WieseEtAl2019} found that software requirements and
management is the software engineering discipline that most hurts scientific
developers, accounting for 23\% of the technical problems reported by study
participants.  The lack of support for requirements is likely due to the
perception that up-front requirements are impossible for research software
\citep{CarverEtAl2007,SegalAndMorris2008}, but when we drop the insistence on
``up-front'' requirements, allowing the requirements to be written
iteratively and incrementally, requirements are feasible \citep{Smith2016}.

>>>>

As part of filling in the measurement template, we listed the artifacts for each
package, and then categorized them by frequency. We group the artifacts into
common, uncommon, and rare categories in Table~\ref{artifactspresent}. Common
artifacts are in 16 to 24 ($>$63\%) of the software packages. Uncommon artifacts
are in 8 to 15 (30-63\%) and rare artifacts are in 1 to 7 ($<$30\%) of the
packages.  We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that
we observed in LBM repositories to those observed and recommended for research
software in general.

@@

In Table~\ref{Tbl_Guidelines}, each row corresponds to an artifact.  For a given
row, a checkmark in one of the columns means that the corresponding guideline
recommends this artifact.  The last column shows whether the artifact appears in
the measured set of LBM software, either not at all (blank), commonly (C),
uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  This challenge
even exists for the ubiquitous README file.  As illustrated by
\citet{PranaEtAl2018}, the content of README files shows significant variation
between projects.  Although some content is reasonably consistent, with 97\% of
README files containing at least one section describing the `What` of the
repository and 88.5 \% offering some `How` content, other categories are more
variable.  For instance, information on `Contribution`, `Why`, and `Who`,
appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
\citep{PranaEtAl2018}.

The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
popularity of recommending a given artifact, but it does not imply that the most
commonly recommended artifacts are the most important artifacts. Moreover, just
because a guideline does not explicitly recommend an artifact does not mean that
the artifact has no value.  The guideline authors may have excluded it because
it is out of the scope of their recommendations or outside their experience. For
instance, an artifact related to uninstall is only explicitly mentioned by
\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
value.  They may simply feel that uninstall is implied by install, or they may
have never asked themselves about the need for separate uninstall instructions.

@@

The top four packages (ESPREsSo, Ludwig, Palabos, and OpenLB) include most of
the common and uncommon artifacts, but not many of the rare artifacts.  They
collectively have all the common artifacts, except Palabos does not have theory
notes, and OpenLB does not appear to use version control.  The top four packages
have most of the uncommon artifacts. Only one them (Ludwig) is missing design
information, and only one (Palabos) does not have a user manual.  However, for
Palabos there was a broken link on the package website indicating that such an
artifact might exist. The developers later fixed this broken link, but this is
not reflected in our data because it was not present at the time of data
collection (mid-2020). Despite the broken link, Palabos does have a detailed and
informative website. The top four ranked packages do not have many of the rare
artifacts. None of them have any explicit API documentation. Three of them
(ESPREsSo, Ludwig, Palabos) have information on contributing to the project and
two (OpenLB, Palabos) have an FAQ section or forum. One (OpenLB) has
verification and validation notes, and a video guide for the software.

Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
appear in the software development guidelines: i) list of related publications,
ii) performance information and iii) video guides.  The occurrence of the first
two items in this list are likely more a consequence of how we collected our
list of LBM artifacts, rather than being a unique practice of LBM developers. We
highlighted publications and performance information because we were explicitly
looking for both as part of our measurement template. Related
publications and performance information certainly appear in other research
software projects; the practices just are not highlighted in guidelines because
they are implicit in other documentation recommendations, such as theory
documentation and user guides. The observation of a video guide artifact is more
novel.  \citet{Fogel2005} recommends videos for open source projects, and other
projects likely use them, but the practice has apparently not yet made it into
research software community's guidelines.  Putting aside the debate of the
effectiveness of learning from print versus video, video documentation is
growing in popularity, with a majority of Generation Z learners preferring video
to print \citep{Genota2018}.

As Table~\ref{Tbl_Guidelines} shows, the LBM community participates in most
practices we found listed in the general research software guidelines; however,
we did not observe, or rarely observed, some recommended practices. For
instance, we rarely saw API documentation for LBM software, but it is a
frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
vanGompelEtAl2016, OrvizEtAl2017, SSI2022, Zadka2018}.  In addition to the items
in the last column of Table~\ref{artifactspresent}, we can add the following
community recommended items that we rarely, if ever, observed:

@@

\item \textbf{A roadmap} is considered by some to be a critical document for an
organization to map out the vision and direction for a product offering
\citep{MunchEtAl2019}.  Roadmaps take different forms, but they all aim to
address the following: \begin{inparaenum}[i)]
	\item Where are we now?,
	\item Where are we going?, and
	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}.
Developers can create a roadmap by the following steps:
\begin{inparaenum}[i)]
	\item define and outline a strategic mission and product vision, 
	\item scan the environment, 
	\item revise and distill the product vision to write the product roadmap, and 
	\item estimate the product life cycle and evaluate the mix of planned
development efforts \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
roadmap has the benefits of providing continuity of purpose, facilitating
stakeholder collaboration and assisting with prioritization \citep{Pichler2012}.
Although we saw some elements of a roadmap in the LBM projects, we did not
observe a specific artifact devoted to this purpose.

\item \textbf{A code of conduct} documents expectations and standards of ethical
behaviour and explicitly states how developers should treat one another
\citep{TouraniEtAl2017}. A code of conduct outlines rules for communication,
establishes enforcement mechanisms for violations and codifies the spirit of a
community, such that anyone can contribute comfortably regardless of, for
example, ethnicity, gender, or sexual orientation \citep{TouraniEtAl2017}. The
top three most popular code's of conduct from \citet{TouraniEtAl2017} are:
\href{https://ubuntu.com/community/code-of-conduct} {Ubuntu Code of Conduct},
\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
{Contributor Covenant}, and \href{https://www.djangoproject.com/conduct/}
{Django Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a
code of conduct on the participation of women. Improved inclusivity means a
wider pool of potential contributors. Moreover, a code of conduct facilitates
meeting ethical obligations for developers that abide by a code of ethics, such
as the IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of
Ontario code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did
not observe any codes of conduct, possibly because the number of developers for
each project is small and because the idea has not yet occurred to them.

@@

\item \textbf{Checklists} can be used to ensure that developers follow best
practices.  Some examples include checklists merging branches into master
\citep{Brown2015}, checklists for saving and sharing changes to the project
\citep{WilsonEtAl2016}, checklists for new and departing team members
\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
releases \citep{HerouxEtAl2008} and checklists for overall software quality
\citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
managing releases, but we did not see any other examples.

\item \textbf{Uninstall instructions} tell a user how to remove the software and
associated libraries from their system.  Uninstall removes files that consume
space and potentially pollute the user's build system with potential future
compilation and linking conflicts.  Only one software package (pyLBM) has
uninstallation instructions.

\item \textbf{API documentation} shows programmers the services or data provided
by the software application (or library) through such resources as its methods
or objects \citep{MengEtAl2018}.  API documentation is a critical factor for API
understandability \citep{MengEtAl2018} (Section~\ref{softwarequalities}). Some
API documentation can be generated using tools like Doxygen, but when learning a
new API developers mostly use code examples and tutorials \citep{MengEtAl2018}.
API documentation is rare within our sample of LBM packages.

@@

Where data is available, LBM software artifacts can be compared to other
research software domains.  Compared to ocean modelling software
\citep{JungEtAl2022}, LBM software seems to pay more attention to artifacts
related to testing.  Compared to Medical Imaging Software \citep{Dong2021}, the
artifact frequency is similar, except LBM software is more likely to have
developer related artifacts, like Contributing, Dependency list, and Design
documentation.

Table~\ref{Tbl_Guidelines} shows that many LBM projects fall short of
recommended best practices for research software.  However, LBM software is not
alone in this.  Many, if not most, research projects fall short of best
practices.  A gap exists in research software development practices and software
engineering recommendations \citep{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
\citet{JohansonAndHasselbring2018} observe that the state-of-the-practice for
research software in industry and academia does not incorporate state-of-the-art
SE tools and methods.  This causes sustainability and reliability problems
\citep{FaulkEtAl2009}. Rather than benefit from capturing and reusing previous
knowledge, projects waste time and energy ``reinventing the wheel''
\citep{deSouzaEtAl2019}.

Software requirements documentation provides a good example of the common
deviation between recommendations and practice.  Although some guidelines
recommend requirements documentation \citep{TobiasEtAl2018, HerouxEtAl2008,
SmithAndKoothoor2016}, in practice research software developers often do not
produce a proper requirements specification \citep{HeatonAndCarver2015}.
\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
found that none of the scientists created requirements specifications, unless
regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
shows requirements are the least commonly produced type of documentation for
research software in general. When looking at the pain points for research
software developers, \citet{WieseEtAl2019} state that software requirements and
management is the software engineering discipline that most hurts scientific
developers, accounting for 23\% of the technical problems reported by study
participants.  The lack of support for requirements is likely due to the
perception that up-front requirements are impossible for research software
\citep{CarverEtAl2007, SegalAndMorris2008}, but when we drop the insistence on
``up-front'' requirements, allowing the requirements to be written
iteratively and incrementally, requirements are feasible \citep{Smith2016}.
